Explicit B-spline regularization in diffeomorphic image registration

### 2013-12-08 paper.tex
ENH:  Final edits?

Free-form deformation (FFD) image registration, characterized by
regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity, good performance, and
{\color{blue} guaranteed parametric continuity}.  Current
research was preceded by related work for geometric modeling
\citep{sederberg1986} and originated with such important contributions
as \cite{szeliski1997,thevenaz1998,rueckert1999}.  Continued
development within this early spline-based paradigm produced
additional innovations such as integrated similarity metrics
\citep[e.g.][]{mattes2003}, additional transformation constraints
\citep[e.g.][]{rohlfing2003}, and notable open source implementations
\citep[e.g.][]{ibanez2005,klein2010,modat2010,shackleford2010}.

@@

It is interesting to note that within this context, traditional FFD
algorithms can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space of
regularized fields using Gaussian convolution, gradient fields are
projected to a smooth space characterized by the B-spline basis
functions.  This analogy was hinted at in our earlier work
\citep{tustison2009} where we showed that fitting the update field to
a B-spline object using a fast approximation routine
\citep{tustison2006} is equivalent to a preconditioning of the
standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect
induced by the ill-conditioned nature of the traditional
gradient-based FFD formulation.%
\footnote{
{\color{blue}{This phenomenon is explained in greater detail in \cite{tustison2009}
but, briefly, the distribution of the uniform B-spline basis functions
spanning the image domain produces long ravines in the image registration
energy topography.  These topographical features cause an
inefficient back-and-forth traversal in classic steepest descent optimization, also known as ``hemstitching.''}}
}
We denoted this new FFD variant as
{\it directly manipulated free-form deformation} (DMFFD) and, as part of the
ITKv4 refactoring efforts, has been implemented for use with the new
registration framework%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

@@

As alluded to earlier, B-spline approximation can also be used for
regularizing time-varying vector fields in an analogous fashion as
Gaussian convolution.  In this vein, and similar to
\cite{de-craene2011}, we reported in \cite{tustison2012a,tustison2012}
the use of an $n$-D + time B-spline object to represent the
characteristic velocity fields.  However, we use the directly
manipulated free-form deformation formulation to improve the solution
convergence.  This also facilitates modeling temporal periodicity and
the enforcement of stationary boundaries.  Both this work and our
earlier work \citep{tustison2009} demonstrate that the DMFFD framework
is potentially applicable to the entire gamut of diffeomorphic
registration algorithms and provides {\color{blue} alternative smoothing 
possibilities with different continuity properties (e.g. $C^2$
versus $C^3$)}.

@@

Given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is given by
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation, spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient at voxel $c$.%
\footnote{
{\color{blue}{For comparison, FFD image registration uses the
calculated gradient which, for a single control point is evaluated using (in 
discretized form assuming no other explicit regularization):
\begin{align}
\label{eq:ffd}
  \delta v_{i_1,\ldots,i_d} = \sum_{c=1}^{N_{\Omega}} 
  \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c 
  \prod_{j=1}^d B_{i_j}(x_j^c).
\end{align}
Intuitively, this can be understood as a distribution of the spatial similarity metric
gradient to the set of control points according to the basis function weighting which, due to
the local property of B-splines, is only non-zero in a localized region surrounding each 
sample point.  The problematic issue with standard FFD image registration is that 
the parametric domain of 
the peripheral B-spline basis functions extends beyond the image boundaries resulting in
a relatively lower weighting contribution effect.  As a corrective, Equation (\ref{eq:dmffd})
normalizes the control point gradient contribution by the actual B-spline weighting overlap
with the image domain.

@@

Based on the needs of one of our colleagues, the first author (N.T.) implemented and
generalized the work of \cite{lee1997}  which was reported in \cite{tustison2006} and
was N.T.'s very first ITK contribution.  
A couple years later the second author (B.A.) requested an FFD implementation for 
ANTs which N.T. naively assumed to be equivalent to applying \cite{tustison2006}
to smoothing the metric gradient (a la Demons).  The subsequent realization 
that this assumption was incorrect and figuring out why the algorithm still worked 
despite it not being a correct FFD implementation resulted in \cite{tustison2009}.
}}
}

@@

Three diffeomorphic algorithms described previously utilizing Gaussian convolution smoothing and their DMFFD counterparts are available through the following classes in the ITK repository: 
\begin{itemize}
  \item {\bf LDDMM:}
    \begin{itemize}
      \item {\tt itk::TimeVaryingVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::TimeVaryingVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::GaussianSmoothingOnUpdateTimeVaryingVelocityFieldTransform}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingBSplineVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf DARTEL:}
    \begin{itemize}
       \item{\tt itk::ConstantVelocityFieldTransform}
       \item{\tt itk::ConstantVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::GaussianExponentialDiffeomorphicTransform} 
      \item {\tt itk::GaussianExponentialDiffeomorphicTransformParametersAdaptor}
      \item {\tt itk::BSplineExponentialDiffeomorphicTransform} 
      \item {\tt itk::BSplineExponentialDiffeomorphicTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf SyN:}
    \begin{itemize}
      \item {\tt itk::SyNImageRegistrationMethod} 
      \item {\tt itk::BSplineSyNImageRegistrationMethod} 
    \end{itemize}  
\end{itemize}
These and other classes (e.g. similarity metrics, optimization methods, and
utility classes) were developed as part of the ITKv4 registration
framework refactoring.  {\color{blue}{Much of the original ITK image registration infrastructure 
was left intact including the so-called ``sparse'' transforms such as various
rigid (versor, Euclidean) and other linear transforms.  The transform classes
contributed by our group were, including those listed above, were meant to 
augment what already existed.  All the transforms listed above are derived from
the {\tt itk::DisplacementFieldTransform} class which permits specification of 
a transform described by a sampled displacement field.  The derived classes are
then modified according to the different transform constraints. Other types
of classes are used to coordinate the image registration process.  The {\tt itk::ImageRegistrationMethodv4} is the base interface for performing all 
image registration steps.  Smoothing and resampling for multi-resolution
image registration is performed in this class as is the calling of the
selected optimizer.  Output consists of a single optimized transform.
Multiple instantiations of this class in series, in conjunction with the 
{\tt itk::CompositeTransform} class, are used to optimize a composition of transforms.
Some of the diffeomorphic approaches do not easily fit into this
generalized optimization framework necessitating specialized method classes such as
those listed above.  The adaptors%
\footnote{
{\color{blue}{We gratefully acknowledge Marius Staring's contribution in providing the broad
design of the adaptors.}}
}
are used to modify the parameters between resolution levels during the course
of transform optimization within the methods classes.  For example, the 
resolution of the displacement field transforms follows that of the image 
resolution and the updating is handled by the corresponding transform adaptors.
}}
Further details can be found in the documentation
provided within the classes themselves.%
\footnote{
http://www.itk.org/Doxygen/html/classes.html
}

@@

antsRegistration --dimensionality 3 \
                 --output ${prefix} \
                 --use-histogram-matching 1 \
                 --initial-moving-transform [${fixed},${moving},1] \
                 --transform Rigid[0.1] \
                 --metric MI[${fixed},${moving},1,32,Regular,0.25] \ 
                 --convergence 1000x500x250 \
                 --smoothing-sigmas 2.0x1.0x0.0mm \
                 --shrink-factors 4x3x2 \
                 --transform Affine[0.1] \
                 --metric MI[${fixed},${moving},1,32,Regular,0.25] \ 
                 --convergence 1000x500x250 \
                 --smoothing-sigmas 2.0x1.0x0.0mm \
                 --shrink-factors 4x3x2 \
                 --transform BSplineSyN[0.1,26,0,3] \
                 --metric CC[${fixed},${moving},1,4] \ 
                 --convergence 100x70x50x20 \
                 --smoothing-sigmas 2x1.0x0.5x0.0mm \
                 --shrink-factors 6x4x2x1

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass (although alignment based on other features is possible).%
\footnote{
{\color{blue}{This step corresponds to the {\tt --initial-moving-transform} option.  Generally, the
user can specify an ITK transform for initialization or can perform an initial
translation based on either the geometric center of the images, the center of the
image intensities, or the origin of the images.  See the help menu `{\tt antsRegistration --help}' for more details.
}}
} 
The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Any pair of images can be specified per metric per stage.%
\footnote{
{
\color{blue}{
To help the reader who wishes to explore {\tt antsRegistration} with specific use of the
parameters used for this study and specified above, we have created a script with a simplified interface and placed it in the {\tt Scripts/} subdirectory of the ANTs repository.  This script, called {\tt antsRegistrationSyN.sh} takes a fixed and moving image and performs a comprehensive (i.e. rigid $\rightarrow$ affine $\rightarrow$ deformable) registration using SyN (`{\tt -t s}') or B-spline SyN (`{\tt -t b}').
}}
}

@@

The number of subjects per cohort is provided in the denotation.
{\color{blue}{Table 1 summarizes core information about the data sets used.}}
Further details of these first four labeled brain data (e.g. labeling protocol,
data sources) are given in \citep{klein2009}.
We also include the labeled brain data provided at the
MICCAI 2012 Grand Challenge and Workshop on Multi-Atlas
Labeling%
\footnote{
https://masi.vuse.vanderbilt.edu/workshop2012
} 
which we denote as MAL35.  This T1-weighted
MRI data set consists of 35 subject MRIs taken from the
Oasis database.%
\footnote{
http://www.oasis-brains.org
}  The corresponding labels were provided by Neuromorphometrics,
Inc.%
\footnote{
http://Neuromorphometrics.com/
}
under academic subscription.

\begin{table*}[htb]
  \caption{
    {\color{blue}{ Brief overview of the data sets used for the SyN vs. B-spline SyN comparison.}}
    }
  \centering
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l c c  c l}
  \toprule
  \multicolumn{1}{c}{Cohort} & Resolution & Number of labels & \multicolumn{1}{c}{Preprocessing} \\
  \midrule
  CUMC12 & $[0.86,0.86,1.5]$ & 131 & \parbox[t]{6.5cm}{rotated to ``cardinal'' pose}\\ 
  IBSR18 & $[1,1.5,1]$ & 96 & \parbox[t]{6.5cm}{aligned to Talaraich, bias corr.}\\ 
  LPBA40 & $[0.86,1.5,0.86]$ & 56  & \parbox[t]{6.5cm}{registered to MNI305, bias corr.}\\
  MGH10 & $[1,1,1]$ & 106 & \parbox[t]{6.5cm}{affine-registered to MNI152, bias corr.}\\ 
  MAL35 & $[1,1,1]$ & 138 & \parbox[t]{6.5cm}{average of 3-4 acquisitions, bias corr.}\\
  \bottomrule
  \end{tabular*}
  \label{table:demo}
\end{table*}

@@

Comparative evaluation of the two SyN registration approaches was performed 
within each cohort using a ``pseudo-geodesic'' approach.  Instead of
registering every subject to every other subject within a data set,
we generated the transforms from each subject to a cohort-specific
shape/intensity template.  {\color{blue}{Not only does this reduce the computational
time required for finding the pairwise transforms between subjects but 
prior work has demonstrated  improvement in registration with this approach\
over direct pairwise registration \citep{klein2010a}.  Since the two algorithms
have been implemented within the same framework, all registration parameters are 
identical (i.e. linear registration stage parameters, winsorizing values, etc.) 
except for the parameters governing the smoothing of the gradient field. 
}}

@@

Additionally, we refined the labelings for each subject of each cohort
using the multi-atlas label fusion algorithm (MALF) developed by \cite{wang2012}
which is also distributed with ANTs.
For a given subject within a data set, every other subject was mapped to 
that subject using the pseudo-geodesic transform.  The set of transformed labelings 
were then used to determine a consensus labeling for that subject.
This was to minimize the obvious {\it observer dimensionality artifacts} where manual
raters observe and label in a single dimension at a time.  This
is most easily seen in the axial or sagittal views of the different cohorts 
as labelings were done primarily in the coronal view (see Figure \ref{fig:labels}).  
We include both sets of results.  This provides two sets of labels per subject
for evaluation.%
\footnote{
{\color{blue}{The MALF labels are available at http://figshare.com/account/projects/196.}}
}

@@

As mentioned previously, a template was constructed for each data
set (cf Figure \ref{fig:templates}) from all cohort images.  Subsequently, 
each image was registered to its corresponding template using either 
SyN or B-spline SyN as described previously (prior linear registration 
stages were identical between the two algorithms).  As a brute-force parameter
exploration is not a part of this work, we rely on previously
reported research \citep{klein2009,avants2011} and our own experience as
authors/developers of the algorithm/software to select parameters which 
demonstrate robust performance across data sets.  For both algorithms,
the gradient step was 0.1 for each of the four multi-resolution levels
with shrink factors of $\{6,4,2,1\}$ and Gaussian smoothing for each
of those levels being $\mathcal{N}\left(0,\{9,4,1,0\}\right)$ in terms
of voxels.  The number of iterations per level were $\{100,100,70,20\}$
with a convergence threshold of $10^{-9}$ and window size of 15 iterations.%
\footnote{
{\color{blue}{
As part of the ITKv4 refactoring, we developed several convergence
monitoring classes (cf the base class {\tt itk::ConvergenceMonitoringFunction}).
Instead of testing for convergence between two successive time points, we use a 
windowed monitoring function which keeps track of a sequence, or window, of 
energy values and determines the convergence from the slope of a line fitted to 
the series.
}}
}

{\color{blue}{
All processing was performed using the linux cluster at the 
University of Virginia%
\footnote{
http://www.uvacse.virginia.edu
}
using the PBS Pro queuing system for managing resources.
}}
The perl scripts used to create the jobs for the cluster are
included in the github account associated with this evaluation.  
For the data sets used in this study, times for B-spline SyN were approximately 
15-40\% greater than Gaussian-based SyN using single-threading and a dense
metric gradient sampling.  
{\color{blue}{
Timing data for specific data sets are given in Table \ref{table:timing}
Timing includes both rigid and affine transform  optimizations.  
\begin{table*}[htb]
  \caption{
    {\color{blue}{Timing (in hours) per registration for both SyN algorithms across data sets.}}
    }
  \centering
  \begin{tabular*}{0.5\textwidth}{@{\extracolsep{\fill}} l c c}
  \toprule
  \multicolumn{1}{c}{Cohort} & SyN & B-spline SyN \\
  \midrule
  CUMC12 & $8.6  \pm  2.3$ & $9.6 \pm 2.1$ \\ 
  IBSR18 & $8.3  \pm  2.7$ & $10.7 \pm 2.8$ \\
  LPBA40 & $6.6 \pm 0.7$ & $9.5 \pm 1.4$ \\
  MGH10 & $4.7  \pm  1.5$ & $5.8 \pm 1.2$ \\
  MAL35 & $10.6 \pm 1.9$ & $14.2 \pm 2.0$ \\
  \bottomrule
  \end{tabular*}
  \label{table:timing}
\end{table*}
}}

>>>>

Free-form deformation (FFD) image registration, characterized by
regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity, good performance, and
guaranteed parametric continuity.  Current
research was preceded by related work for geometric modeling
\citep{sederberg1986} and originated with such important contributions
as \cite{szeliski1997,thevenaz1998,rueckert1999}.  Continued
development within this early spline-based paradigm produced
additional innovations such as integrated similarity metrics
\citep[e.g.][]{mattes2003}, additional transformation constraints
\citep[e.g.][]{rohlfing2003}, and notable open source implementations
\citep[e.g.][]{ibanez2005,klein2010,modat2010,shackleford2010}.

@@

It is interesting to note that within this context, traditional FFD
algorithms can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space of
regularized fields using Gaussian convolution, gradient fields are
projected to a smooth space characterized by the B-spline basis
functions.  This analogy was hinted at in our earlier work
\citep{tustison2009} where we showed that fitting the update field to
a B-spline object using a fast approximation routine
\citep{tustison2006} is equivalent to a preconditioning of the
standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect
induced by the ill-conditioned nature of the traditional
gradient-based FFD formulation.%
\footnote{
This phenomenon is explained in greater detail in \cite{tustison2009}
but, briefly, the distribution of the uniform B-spline basis functions
spanning the image domain produces long ravines in the image registration
energy topography.  These topographical features cause an
inefficient back-and-forth traversal in classic steepest descent optimization, also known as ``hemstitching.''
}
We denoted this new FFD variant as
{\it directly manipulated free-form deformation} (DMFFD) and, as part of the
ITKv4 refactoring efforts, has been implemented for use with the new
registration framework%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

@@

As alluded to earlier, B-spline approximation can also be used for
regularizing time-varying vector fields in an analogous fashion as
Gaussian convolution.  In this vein, and similar to
\cite{de-craene2011}, we reported in \cite{tustison2012a,tustison2012}
the use of an $n$-D + time B-spline object to represent the
characteristic velocity fields.  However, we use the directly
manipulated free-form deformation formulation to improve the solution
convergence.  This also facilitates modeling temporal periodicity and
the enforcement of stationary boundaries.  Both this work and our
earlier work \citep{tustison2009} demonstrate that the DMFFD framework
is potentially applicable to the entire gamut of diffeomorphic
registration algorithms and provides alternative smoothing 
possibilities with different continuity properties (e.g. $C^2$
versus $C^3$).

@@

Given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is given by
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation, spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient at voxel $c$.%
\footnote{
For comparison, FFD image registration uses the
calculated gradient which, for a single control point is evaluated using (in 
discretized form assuming no other explicit regularization):
\begin{align}
\label{eq:ffd}
  \delta v_{i_1,\ldots,i_d} = \sum_{c=1}^{N_{\Omega}} 
  \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c 
  \prod_{j=1}^d B_{i_j}(x_j^c).
\end{align}
Intuitively, this can be understood as a distribution of the spatial similarity metric
gradient to the set of control points according to the basis function weighting which, due to
the local property of B-splines, is only non-zero in a localized region surrounding each 
sample point.  The problematic issue with standard FFD image registration is that 
the parametric domain of 
the peripheral B-spline basis functions extends beyond the image boundaries resulting in
a relatively lower weighting contribution effect.  As a corrective, Equation (\ref{eq:dmffd})
normalizes the control point gradient contribution by the actual B-spline weighting overlap
with the image domain.

@@

Based on the needs of one of our colleagues, the first author (N.T.) implemented and
generalized the work of \cite{lee1997}  which was reported in \cite{tustison2006} and
was N.T.'s very first ITK contribution.  
A couple years later the second author (B.A.) requested an FFD implementation for 
ANTs which N.T. naively assumed to be equivalent to applying \cite{tustison2006}
to smoothing the metric gradient (a la Demons).  The subsequent realization 
that this assumption was incorrect and figuring out why the algorithm still worked 
despite it not being a ``correct'' FFD implementation resulted in \cite{tustison2009}.
}

@@

Three diffeomorphic algorithms described previously utilizing Gaussian convolution smoothing and their DMFFD counterparts are available through the following classes in the ITK repository: 
\begin{itemize}
  \item {\bf LDDMM:}
    \begin{itemize}
      \item {\tt itk::TimeVaryingVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::TimeVaryingVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::GaussianSmoothingOnUpdateTimeVaryingVelocityFieldTransform}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingBSplineVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf DARTEL:}
    \begin{itemize}
       \item{\tt itk::ConstantVelocityFieldTransform}
       \item{\tt itk::ConstantVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::GaussianExponentialDiffeomorphicTransform} 
      \item {\tt itk::GaussianExponentialDiffeomorphicTransformParametersAdaptor}
      \item {\tt itk::BSplineExponentialDiffeomorphicTransform} 
      \item {\tt itk::BSplineExponentialDiffeomorphicTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf SyN:}
    \begin{itemize}
      \item {\tt itk::SyNImageRegistrationMethod} 
      \item {\tt itk::BSplineSyNImageRegistrationMethod} 
    \end{itemize}  
\end{itemize}
These and other classes (e.g. similarity metrics, optimization methods, and
utility classes) were developed as part of the ITKv4 registration
framework refactoring.  Much of the original ITK image registration infrastructure 
was left intact including the so-called ``sparse'' transforms such as various
rigid (versor, Euclidean) and other linear transforms.  The transform classes
contributed by our group were, including those listed above, were meant to 
augment what already existed.  All the transforms listed above are derived from
the {\tt itk::DisplacementFieldTransform} class which permits specification of 
a transform described by a sampled displacement field.  The derived classes are
then modified according to the different transform constraints. Other types
of classes are used to coordinate the image registration process.  The {\tt itk::ImageRegistrationMethodv4} is the base interface for performing all 
image registration steps.  Smoothing and resampling for multi-resolution
image registration is performed in this class as is the calling of the
selected optimizer.  Output consists of a single optimized transform.
Multiple instantiations of this class in series, in conjunction with the 
{\tt itk::CompositeTransform} class, are used to optimize a composition of transforms.
Some of the diffeomorphic approaches do not easily fit into this
generalized optimization framework necessitating specialized method classes such as
those listed above.  The adaptors%
\footnote{
We gratefully acknowledge Marius Staring's contribution in providing the broad
design of the adaptors.
}
are used to modify the parameters between resolution levels during the course
of transform optimization within the methods classes.  For example, the 
resolution of the displacement field transforms follows that of the image 
resolution and the updating is handled by the corresponding transform adaptors.
Further details can be found in the documentation
provided within the classes themselves.%
\footnote{
http://www.itk.org/Doxygen/html/classes.html
}

@@

antsRegistration --dimensionality 3 \
                 --output ${prefix} \
                 --use-histogram-matching 1 \
                 --initial-moving-transform [${fixed},${moving},1] \
                 --transform Rigid[0.1] \
                 --metric MI[${fixed},${moving},1,32,Regular,0.25] \ 
                 --convergence 1000x500x250x100 \
                 --smoothing-sigmas 3x2x1x0 \
                 --shrink-factors 8x4x2x1 \
                 --transform Affine[0.1] \
                 --metric MI[${fixed},${moving},1,32,Regular,0.25] \ 
                 --convergence 1000x500x250x100 \
                 --smoothing-sigmas 3x2x1x0 \
                 --shrink-factors 8x4x2x1 \
                 --transform BSplineSyN[0.1,26,0,3] \
                 --metric CC[${fixed},${moving},1,4] \ 
                 --convergence 100x70x50x20 \
                 --smoothing-sigmas 3x2x1x0 \
                 --shrink-factors 6x4x2x1

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass (although alignment based on other features is possible).%
\footnote{
This step corresponds to the {\tt --initial-moving-transform} option.  Generally, the
user can specify an ITK transform for initialization or can perform an initial
translation based on either the geometric center of the images, the center of the
image intensities, or the origin of the images.  See the help menu `{\tt antsRegistration --help}' for more details.
} 
The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Any pair of images can be specified per metric per stage.%
\footnote{
To help the reader who wishes to explore {\tt antsRegistration} with specific use of the
parameters used for this study and specified above, we have created a script with a simplified interface and placed it in the {\tt Scripts/} subdirectory of the ANTs repository.  This script, called {\tt antsRegistrationSyN.sh} takes a fixed and moving image and performs a comprehensive (i.e. rigid $\rightarrow$ affine $\rightarrow$ deformable) registration using SyN (`{\tt -t s}') or B-spline SyN (`{\tt -t b}').
}

@@

The number of subjects per cohort is provided in the denotation.
Table 1 summarizes core information about the data sets used.
Further details of these first four labeled brain data (e.g. labeling protocol,
data sources) are given in \citep{klein2009}.
We also include the labeled brain data provided at the
MICCAI 2012 Grand Challenge and Workshop on Multi-Atlas
Labeling%
\footnote{
https://masi.vuse.vanderbilt.edu/workshop2012
} 
which we denote as MAL35.  This T1-weighted
MRI data set consists of 35 subject MRIs taken from the
Oasis database.%
\footnote{
http://www.oasis-brains.org
}  The corresponding labels were provided by Neuromorphometrics,
Inc.%
\footnote{
http://Neuromorphometrics.com/
}
under academic subscription.

\begin{table*}[htb]
  \caption{
    Brief overview of the data sets used for the SyN vs. B-spline SyN comparison.
    }
  \centering
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l c c  c l}
  \toprule
  \multicolumn{1}{c}{Cohort} & Resolution & Number of labels & \multicolumn{1}{c}{Preprocessing} \\
  \midrule
  CUMC12 & $[0.86,0.86,1.5]$ & 131 & \parbox[t]{6.5cm}{rotated to ``cardinal'' pose}\\ 
  IBSR18 & $[1,1.5,1]$ & 96 & \parbox[t]{6.5cm}{aligned to Talaraich, bias corr.}\\ 
  LPBA40 & $[0.86,1.5,0.86]$ & 56  & \parbox[t]{6.5cm}{registered to MNI305, bias corr.}\\
  MGH10 & $[1,1,1]$ & 106 & \parbox[t]{6.5cm}{affine-registered to MNI152, bias corr.}\\ 
  MAL35 & $[1,1,1]$ & 138 & \parbox[t]{6.5cm}{average of 3-4 acquisitions, bias corr.}\\
  \bottomrule
  \end{tabular*}
  \label{table:demo}
\end{table*}

@@

Comparative evaluation of the two SyN registration approaches was performed 
within each cohort using a ``pseudo-geodesic'' approach.  Instead of
registering every subject to every other subject within a data set,
we generated the transforms from each subject to a cohort-specific
shape/intensity template.  Not only does this reduce the computational
time required for finding the pairwise transforms between subjects but 
prior work has demonstrated  improvement in registration with this approach\
over direct pairwise registration \citep{klein2010a}.  Since the two algorithms
have been implemented within the same framework, all registration parameters are 
identical (i.e. linear registration stage parameters, winsorizing values, etc.) 
except for the parameters governing the smoothing of the gradient field.

@@

Additionally, we refined the labelings for each subject of each cohort
using the multi-atlas label fusion algorithm (MALF) developed by \cite{wang2012}
which is also distributed with ANTs.
For a given subject within a data set, every other subject was mapped to 
that subject using the pseudo-geodesic transform.  The set of transformed labelings 
were then used to determine a consensus labeling for that subject.
This was to minimize the obvious {\it observer dimensionality artifacts} where manual
raters observe and label in a single dimension at a time.  This
is most easily seen in the axial or sagittal views of the different cohorts 
as labelings were done primarily in the coronal view (see Figure \ref{fig:labels}).  
We include both sets of results.  This provides two sets of labels per subject
for evaluation.%
\footnote{
The MALF labels are available at http://figshare.com/account/projects/196.
}

@@

As mentioned previously, a template was constructed for each data
set (cf Figure \ref{fig:templates}) from all cohort images.  Subsequently, 
each image was registered to its corresponding template using either 
SyN or B-spline SyN as described previously (prior linear registration 
stages were identical between the two algorithms).  As a brute-force parameter
exploration is not a part of this work, we rely on previously
reported research \citep{klein2009,avants2011} and our own experience as
authors/developers of the algorithm/software to select parameters which 
demonstrate robust performance across data sets.  For both algorithms,
the gradient step was 0.1 for each of the four multi-resolution levels
with shrink factors of $\{6,4,2,1\}$ and Gaussian smoothing for each
of those levels being $\mathcal{N}\left(0,\{9,4,1,0\}\right)$ in terms
of voxels.  The number of iterations per level were $\{100,100,70,20\}$
with a convergence threshold of $10^{-9}$ and window size of 15 iterations.%
\footnote{
As part of the ITKv4 refactoring, we developed several convergence
monitoring classes (cf the base class {\tt itk::ConvergenceMonitoringFunction}).
Instead of testing for convergence between two successive time points, we use a 
windowed monitoring function which keeps track of a sequence, or window, of 
energy values and determines the convergence from the slope of a line fitted to 
the series.
}

All processing was performed using the linux cluster at the 
University of Virginia%
\footnote{
http://www.uvacse.virginia.edu
}
using the PBS Pro queuing system for managing resources.
The perl scripts used to create the jobs for the cluster are
included in the github account associated with this evaluation.  
For the data sets used in this study, times for B-spline SyN were approximately 
15-40\% greater than Gaussian-based SyN using single-threading and a dense
metric gradient sampling.  
Timing data for specific data sets are given in Table \ref{table:timing}
Timing includes both rigid and affine transform  optimizations.  
\begin{table*}[htb]
  \caption{
    Timing (in hours) per registration for both SyN algorithms across data sets.
    }
  \centering
  \begin{tabular*}{0.5\textwidth}{@{\extracolsep{\fill}} l c c}
  \toprule
  \multicolumn{1}{c}{Cohort} & SyN & B-spline SyN \\
  \midrule
  CUMC12 & $8.6  \pm  2.3$ & $9.6 \pm 2.1$ \\ 
  IBSR18 & $8.3  \pm  2.7$ & $10.7 \pm 2.8$ \\
  LPBA40 & $6.6 \pm 0.7$ & $9.5 \pm 1.4$ \\
  MGH10 & $4.7  \pm  1.5$ & $5.8 \pm 1.2$ \\
  MAL35 & $10.6 \pm 1.9$ & $14.2 \pm 2.0$ \\
  \bottomrule
  \end{tabular*}
  \label{table:timing}
\end{table*}
